Pangenome graph layout by HOGWILD! Path-Guided Stochastic Gradient Descent

### 2023-06-20 manuscript/main.tex
abstract and introduction

\abstract{
\textbf{Motivation:}
The availability of many high-quality genomes has led to the need for models that allow the study of genomic variability within entire populations.
Pangenome graphs can represent the full diversity between multiple genomes, but they may exhibit complex structures due to common patterns of genome variation.
These nonlinear structures introduce difficulty in downstream analyses, visualization, and interpretation. \\
\textbf{Results:}
To address these challenges, we propose a novel layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
Our approach arranges the nodes of a pangenome graph by leveraging the HOGWILD! strategy and it is guided by the path position information of the genomes represented in the graph.
We demonstrate that our implementation can efficiently compute the latent structure of gigabase-scale pangenome graphs, exposing their biological features. \\
\textbf{Availability:}
The PG-SGD algorithm is implemented in \textit{odgi} which is published as free software under the MIT open source license.
Source code is available at \url{https://github.com/pangenome/odgi} and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}.
\textit{odgi} can be installed via Bioconda \url{https://bioconda.github.io/recipes/odgi/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/odgi.scm}. \\
\textbf{Contact:} \href{simon.heumos@qbic.uni-tuebingen.de}{simon.heumos@qbic.uni-tuebingen.de},
\href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
}

@@

\section{Background}
A pangenome1 models the entire set of genomic elements of a given population. It can efficiently be encoded2 in the form of a variation graph, which embeds the linear sequences of the pangenome as paths in the graphs themselves. 
\\
In pangenome graphs models, DNA sequences are incorporated as nodes with edges connecting the nodes as they occur as sequences representing the graph.
\\
Pangenome graphs \cite{computational2016computational}. blubb.
\\
variation graph model
\\
SGD-layout. Layout in general.
\\ Very light background.
%Pangenome graphs built from raw sets of alignments may have complex structures generated by common patterns of genome variation.

>>>>

\abstract{
\textbf{Motivation:}
The increasing availability of high-quality genomes demands models that allow for the study of genomic variability within entire populations.
Pangenome graphs can represent the full genetic diversity between multiple genomes, but they may exhibit complex structures due to common, nonlinear patterns of genome variation and evolution.
These structures introduce difficulty in downstream analyses, visualization, and interpretation. \\
\textbf{Results:}
To address these challenges, we propose a novel layout algorithm, the Path-Guided Stochastic Gradient Descent (PG-SGD).
Our method, driven by the HOGWILD! strategy, arranges the nodes of a pangenome graph using the path position information of the genomes represented in the graph.
We demonstrate that our implementation can efficiently compute the latent structure of gigabase-scale pangenome graphs, revealing their biological features. \\
\textbf{Availability:}
The PG-SGD algorithm has been integrated in \textit{ODGI} which is released as free software under the MIT open source license.
Source code is available at \url{https://github.com/pangenome/odgi} and documentation at \url{https://odgi.readthedocs.io/en/latest/rst/tutorials/sort_layout.html}.
\textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu}
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
}

@@

\section{Introduction}
Reference genomes are the most widely used resource in genetics, serving as a foundation for a variety of downstream analyses
such as gene annotation, read mapping, and variant detection (\FIXME{XXX WE NEED CITATIONS XXX}).
However, with the availability of thousands of high-quality genomes per species, this linear model has shown its weaknesses.
A single genome can not faithfully represent the genetic diversity of a species, leading to the reference bias problem (\FIXME{XXX WE NEED CITATIONS XXX}).
A pangenome models the entire set of genomic elements of a given population \citep{Tettelin_2008,cpang2018,Eizenga_2020}.
It can be represented by a graph data structure incorporating sequences as nodes and their connections as edges ((\FIXME{XXX A FIGURE MIGHT HELP XXX}).
These nodes are shared for identical sequences, such as homologs, paralogs, and orthologs.
In the variation graph model \citep{Garrison:2018} reads, contigs, haplotypes, or individual chromosomes are represented as paths in the graph.
A bi-directed graph can even contain both strands of DNA as paths. \\

The layout of these pangenome graphs, that is the arrangement of nodes and edges in a representation, provide an accessible representation of genetic variation across multiple genomes.
Graph sorting aims to find the best node order for a layout to minimizes overlapping nodes or edges, reduces edge crossings, and promote intuitive understanding of the data by displaying structure and important features in an understandable way.
\FIXME{
	Existing layout algorithms range from simple, force-directed methods to more complex, hierarchical layouts, each presenting unique strengths and drawbacks.
	Force-directed methods offer an aesthetic layout based on a physics simulation but may struggle with scalability for larger graphs (XXX WE NEED CITATIONS XXX).
	In contrast, hierarchical layouts handle layered relationships and dependencies, yet can exhibit limitations when dealing with non-hierarchical data (XXX WE NEED CITATIONS XXX)
}

With the ever-growing scale and complexity of genomic data, there is a need for advanced layout algorithms that can optimally balance factors such as
efficiency, scalability, readability, and the ability to reveal significant biological patterns.
Pangenome graphs embed linear sequences as paths in the graph, but to our knowledge, no algorithm takes into account this biological information in the sorting.
We present a new layout algorithm to simplify a pangenome graph, by using path-guided stochastic gradient descent (\FIXME{CITATION}) to move a single pair of nodes at a time.
\FIXME{A LITTLE BIT MORE ABOUT THE SGD APPROACH?}
We exemplify how our implementation is a key step in general pangenome analyses such as pangenome graph buildiong and analysis (\FIXME{CITE PGGB, ACROCENTRIC PAPER, and maybe HPRC PAPER}).
